Fast Sorting on Modern Computers

### 2005-08-30 docs/tech_report/elementary.tex
formatting fixes. -- and "\.  "

Knuth also presents the number of changes to the left-to-right minimum in the
best case (0 changes) and worst cases ($N-1$ changes). In the worst case, the
array is sorted in reverse order, and so the left-to-right minimum changes on
every element. Interestingly, the worst cases for the number of changes is not
the worst case for branch mispredictions. If the minimum changes for every
element, then the comparison branch will always resolve in the same direction,
and thus be almost perfectly predictable. From the point of view of branch
prediction, frequent changes (preferably with no simple pattern) are the main
indicator of poor performance, rather than the absolute number of times that the
branch resolves in each direction. The worst case for selection sort would
probably be an array which is partially sorted in reverse order --- the
left-to-right minimum would change frequently, but not predictably so.

>>>>

Knuth also presents the number of changes to the left-to-right minimum in the
best case (0 changes) and worst cases ($N-1$ changes). In the worst case, the
array is sorted in reverse order, and so the left-to-right minimum changes on
every element. Interestingly, the worst cases for the number of changes is not
the worst case for branch mispredictions. If the minimum changes for every
element, then the comparison branch will always resolve in the same direction,
and thus be almost perfectly predictable. From the point of view of branch
prediction, frequent changes (preferably with no simple pattern) are the main
indicator of poor performance, rather than the absolute number of times that the
branch resolves in each direction. The worst case for selection sort would
probably be an array which is partially sorted in reverse order - the
left-to-right minimum would change frequently, but not predictably so.

### 2005-08-30 docs/tech_report/merge.tex
formatting fixes. -- and "\.  "

Figure \ref{mergesorts-1} shows the behaviour of the individual branches within
algorithm N.  The labels in the graphs correspond to the names of the steps in
Knuth's algorithm (we also used the same names for the labels of branch targets
in our C source code. Branch N3 is the main comparison branch; it is used to
choose the smaller of the elements from the start of the two sub-arrays being
merged. The direction of this branch is completely random -- it depends entirely
on the (in this case random) data. Figure \ref{mergesorts-1} also shows that
this branch is almost entirely unpredictable using a bimodal predictor. The
result for a two-level predictor is much the same, because there is no pattern
in the outcome of these branches. Branches `\n{N3 i==j}' (check for the current
merge being complete), `\n{N5}' and `\n{N9} (both checks for the end of a run of
increasing elements) are control-flow branches, and are almost entirely
predictable. Figure \ref{mergesorts-2} shows that the comparison branch in
Algorithm S is just as unpredictable as that in algorithm N. However, given that
the merge in algorithm S is entirely regular (see section \ref{Algorithm S}) it
is possible to eliminate one of the control-flow branches.

Base mergesort is an optimized algorithm that reverses the sorting direction of
every second sub-array, pre-sorts small sub-arrays, and applies a number of
other optimizations. As Figure \ref{mergesorts-3} shows, our base mergesort has
eliminated some of the control-flow branches. The main comparison branch is
split into two separate branches --- one iterating from left-to-right over the
increasing sub-array (`\n{Left}'), and the other from right-to-left over the
decreasing sub-array (`\n{Right}').  However, base mergesort appears to need a
total of 50\% more comparison branches that either Algorithm S or N.

>>>>

Figure \ref{mergesorts-1} shows the behaviour of the individual branches within
algorithm N. The labels in the graphs correspond to the names of the steps in
Knuth's algorithm (we also used the same names for the labels of branch targets
in our C source code. Branch N3 is the main comparison branch; it is used to
choose the smaller of the elements from the start of the two sub-arrays being
merged. The direction of this branch is completely random - it depends entirely
on the (in this case random) data. Figure \ref{mergesorts-1} also shows that
this branch is almost entirely unpredictable using a bimodal predictor. The
result for a two-level predictor is much the same, because there is no pattern
in the outcome of these branches. Branches `\n{N3 i==j}' (check for the current
merge being complete), `\n{N5}' and `\n{N9} (both checks for the end of a run of
increasing elements) are control-flow branches, and are almost entirely
predictable. Figure \ref{mergesorts-2} shows that the comparison branch in
Algorithm S is just as unpredictable as that in algorithm N. However, given that
the merge in algorithm S is entirely regular (see section \ref{Algorithm S}) it
is possible to eliminate one of the control-flow branches.

Base mergesort is an optimized algorithm that reverses the sorting direction of
every second sub-array, pre-sorts small sub-arrays, and applies a number of
other optimizations. As Figure \ref{mergesorts-3} shows, our base mergesort has
eliminated some of the control-flow branches. The main comparison branch is
split into two separate branches - one iterating from left-to-right over the
increasing sub-array (`\n{Left}'), and the other from right-to-left over the
decreasing sub-array (`\n{Right}'). However, base mergesort appears to need a
total of 50\% more comparison branches that either Algorithm S or N.

### 2005-08-30 docs/tech_report/quick.tex
removed a TODO
formatting fixes. -- and "\.  "

%TODO
% \textbf{[David] I'm not sure what to do with this section. Part of the
% problem is that the pictures are very difficult to understand. Another
% problem is that we are trying to show differences that are almost
% invisible. The median of 9 picture should definitely be removed from
% this subsection. However, I suspect that the whole of this subsection should
% go. What do we want to say? One really puzzling result is that there
% seem to be no median branches at all for cache quicksort. This looks
% strikingly like a programming error.}

% \textbf{[David] On thing that's badly needed in this section is
% a detailed description of what these funny charts all mean. However,
% the appropriate place for most of this is the place where the first of
% these charts appeared.}

@@

Overall, base quicksort has a 35\% branch misprediction rate, whereas the rate
for memory-tuned quicksort is only 34\%, while the number of branches also
decreases. The main reason for this is the reduction in branches by not
attempting to move the pivot elements. Attempting to move these elements
usually results in a branch misprediction. The main reason is that most other
elements move at least one position, and thus cause the insertion sort loop to
iterate at least once. But the pivot elements never move -- they are already in
the correct place.

When we look at multi-quicksort, there is a 19\% drop in the number of
iterations of the \n{i} and \n{j} loops, with a corresponding reduction in the
number of mispredictions. Interestingly, the percentage misprediction rate of
the \n{i} and \n{j} branches does not change -- it is 37\% in both cases.

>>>>

@@

Overall, base quicksort has a 35\% branch misprediction rate, whereas the rate
for memory-tuned quicksort is only 34\%, while the number of branches also
decreases. The main reason for this is the reduction in branches by not
attempting to move the pivot elements. Attempting to move these elements usually
results in a branch misprediction. The main reason is that most other elements
move at least one position, and thus cause the insertion sort loop to iterate at
least once. But the pivot elements never move - they are already in the correct
place.

When we look at multi-quicksort, there is a 19\% drop in the number of
iterations of the \n{i} and \n{j} loops, with a corresponding reduction in the
number of mispredictions. Interestingly, the percentage misprediction rate of
the \n{i} and \n{j} branches does not change - it is 37\% in both cases.
